Report for Programming Assignment 1

### 2017-08-31 Programming-Assignment-1/PA-1_report.tex
PA-1 Update 11

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{float}

@@

\begin{document}
\maketitle
\begin{flushleft}
The ``special" functions in consideration for this assignment were:
\begin{gather*}
f_{Q}(x, y) = 1.125x^2 + 0.5xy + 0.75y^2 + 2x + 2y \\
f_{LL}(x, y) = 0.5(x^2 + y^2) + 50\log\left(1 + e^{-0.5y}\right) + 50\log\left(1 + e^{0.2x}\right) \\
f_{H}(x, y) = 0.1(x^2 + y - 11)^2 + 0.1(x + y^2 - 7)^2 \\
f_{R}(x, y) = 0.002(1 - x)^2 + 0.2(y - x^2)^2
\end{gather*}

These functions were plotted using \texttt{GNUplot}, a terminal based plotting tool. The code for plotting can be found in the file: \texttt{plotter.gnuplot}. A few important observations could be made:
\begin{itemize}
\item In the domain of consideration i.e., \([-6, 6]\times[-6, 6]\) for \(f_{Q}, f_{LL} \text{ \& } f_{H}\) and \([-3, 3] \times [-6, 6]\) for \(f_{R}\) --- \(f_{Q}\) and \(f_{LL}\) were found to convex, and \(f_{H}\) and \(f_{R}\) were found to be non-convex. All these inferences were based of the surfaces plotted.
\item \(f_{Q}\) and \(f_{LL}\) have 1 minimum each, whereas \(f_{H}\) has 4 local minima and 1 local maximum. The problem with \(f_{R}\) is the existence of a valley like structure.
\end{itemize}

\begin{figure}[H]
\begin{minipage}{0.45\linewidth}
\centering
\includegraphics[width=0.75\textwidth]{./images/quadratic_function}
\caption{Note the basin (dark region in the middle)}
\end{minipage}
\hfill
\begin{minipage}{0.45\linewidth}
\centering
\includegraphics[width=0.75\textwidth]{./images/ridge_regularized_logistic_regression}
\caption{Note the basin (slightly darker region near \((-4, 4)\))}
\end{minipage}
\end{figure}

\begin{figure}[H]
\begin{minipage}{0.45\linewidth}
\centering
\includegraphics[width=0.75\textwidth]{./images/himmelblaus_function}
\caption{Note the 4 basins and a plateau approximately in the middle of them}
\end{minipage}
\hfill
\begin{minipage}{0.45\linewidth}
\centering
\includegraphics[width=0.75\textwidth]{./images/rosenbrock_banana_function}
\caption{Note the valley (dark parabolic structure)}
\end{minipage}
\end{figure}

@@

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
& 0.3 & 0.1 & 0.01 \\
\hline
\(f_{Q}(x, y)\) & 0 & 0 & 0\\
\hline
\(f_{LL}(x, y)\) & 0 & 0 & 0\\
\hline
\(f_{H}(x, y)\) & 1.769 & 0 & 0 \\
\hline
\(f_{R}(x, y)\) & 0.00092 & 0.00106 & 0.00112 \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\(f_{Q}(x, y)\) & \(f_{LL}(x, y)\) & \(f_{H}(x, y)\) & \(f_{R}(x, y)\) \\
\hline
0 & 0 & 0 & 0.00078 \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\(f_{Q}(x, y)\) & \(f_{LL}(x, y)\) & \(f_{H}(x, y)\) & \(f_{R}(x, y)\) \\
\hline
0 & 0 & 0 & 0.00062 \\
\hline
\end{tabular}
\end{center}

@@

>>>>

\documentclass{article}
\usepackage[margin=0.73in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{hyperref}

@@

\begin{document}
\maketitle

\section*{Instructions for examining the programs}
\begin{itemize}
\item There are three important programs: \texttt{fixed\_step\_size.py} for fixed step size (Question 2), \texttt{backtracking\_ls.py} for backtracing line search (Question 3) and \texttt{inv\_step\_size.py} for learning rate = \(\frac{1}{k}\) (Question 4).
\item All of the above mentioned files depend on \texttt{NumPy} and a module \texttt{functions.py}, included with these programs.
\item To test a program, run \texttt{python3 <name\_of\_the\_program> -h}. The options are there for your convenience.
\item Since analytically arriving at a solution for \(f_{LL}\) was not possible, I wrote a smaller program to find the roots of partial derivatives using the Newton-Raphson Method \texttt{newton\_raphson.py}. 
\item The plots were generated using \texttt{GNUplot}. The plots can be obtained by running \texttt{gnuplot plotter.gnuplot}.
\end{itemize}

\section*{Main Report}
\begin{flushleft}
The ``special" functions in consideration for this assignment were:
\begin{gather*}
f_{Q}(x, y) = 1.125x^2 + 0.5xy + 0.75y^2 + 2x + 2y \\
f_{LL}(x, y) = 0.5(x^2 + y^2) + 50\log\left(1 + e^{-0.5y}\right) + 50\log\left(1 + e^{0.2x}\right) \\
f_{H}(x, y) = 0.1(x^2 + y - 11)^2 + 0.1(x + y^2 - 7)^2 \\
f_{R}(x, y) = 0.002(1 - x)^2 + 0.2(y - x^2)^2
\end{gather*}

\begin{itemize}
\item In the domain of consideration i.e., \([-6, 6]\times[-6, 6]\) for \(f_{Q}, f_{LL} \text{ \& } f_{H}\) and \([-3, 3] \times [-6, 6]\) for \(f_{R}\) --- \(f_{Q}\) and \(f_{LL}\) were found to convex, and \(f_{H}\) and \(f_{R}\) were found to be non-convex. All these inferences were based of the surfaces plotted.
\item \(f_{Q}\) and \(f_{LL}\) have 1 minimum each, whereas \(f_{H}\) has 4 local minima and 1 local maximum. The problem with \(f_{R}\) is the existence of a valley like structure.
\end{itemize}

\begin{figure}[H]
\begin{minipage}{0.45\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{./images/quadratic_function}
\caption{Note the basin (dark region in the middle)}
\end{minipage}
\hfill
\begin{minipage}{0.45\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{./images/ridge_regularized_logistic_regression}
\caption{Note the basin (slightly darker region near \((-4, 4)\))}
\end{minipage}
\end{figure}

\begin{figure}[H]
\begin{minipage}{0.45\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{./images/himmelblaus_function}
\caption{Note the 4 basins and a plateau approximately in the middle of them}
\end{minipage}
\hfill
\begin{minipage}{0.45\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{./images/rosenbrock_banana_function}
\caption{Note the valley (dark parabolic structure)}
\end{minipage}
\end{figure}

@@

\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\(\eta \rightarrow\) & 0.3 & 0.1 & 0.01 \\
\hline
\(f_{Q}(x, y)\) & 0 & 0 & 0\\
\hline
\(f_{LL}(x, y)\) & 0 & 0 & 0\\
\hline
\(f_{H}(x, y)\) & 1.769 & 0 & 0 \\
\hline
\(f_{R}(x, y)\) & 0.00092 & 0.00106 & 0.00112 \\
\hline
\end{tabular}
\caption{\(f(\mathbf{x}^{(1000)}) - f(\mathbf{x}^{*})\) for Fixed Step Size}
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\(f_{Q}(x, y)\) & \(f_{LL}(x, y)\) & \(f_{H}(x, y)\) & \(f_{R}(x, y)\) \\
\hline
0 & 0 & 0 & 0.00078 \\
\hline
\end{tabular}
\caption{\(f(\mathbf{x}^{(1000)}) - f(\mathbf{x}^{*})\) for Backtracking Line Search}
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\(f_{Q}(x, y)\) & \(f_{LL}(x, y)\) & \(f_{H}(x, y)\) & \(f_{R}(x, y)\) \\
\hline
0 & 0 & 0 & 0.00062 \\
\hline
\end{tabular}
\caption{\(f(\mathbf{x}^{(1000)}) - f(\mathbf{x}^{*})\) for Annealing Schedule}
\end{table}

@@

From Yann LeCun's paper \href{http://yann.lecun.com/exdb/publis/pdf/lecun-simard-pearlmutter-93.pdf}{Automatic Learning Rate Maximization by On-Line Estimation of the Hessian's Eigenvectors}, the optimal learning rate seems to be \(\eta_{\text{opt}} = \frac{1}{\lambda_{max}}\), where \(\lambda_{max}\) is the maximum eigenvalue of the Hessian matrix. If the learning rate is too high, then it might be the case that the learning rate set is too greater than the optimal learning rate itself. For example: in the case of Himmelblau's function \(f_{H}(x, y)\), the eigenvalues of the Hessian at \((2, 3)\) are : \(1.338, 10.462\), which means that the optimal learning rate is: \(\approx 0.095\) to start with.
